---
title: "AJP PG Publication Code"
author: "Abby Beatty"
date: "May 29, 2020"
output: html_document
---

##Load Necessary Packages
```{r echo=T, include=T, warning=F, message=F}

library(ggplot2)
library(nlme)
library(multcomp)
library(emmeans)
library(MuMIn)
library(bestNormalize)
library(lmtest)
```

##Summary Function for Mean and specifications for graphing parameters
```{r}
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=TRUE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)
    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=TRUE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }
    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )
    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))
    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult
    return(datac)
}
pd <- position_dodge(0.02)
dodge <- position_dodge(width = 0.6)
```

##READ IN AND CALCULATE AVERAGES
```{R}
Q=read.csv("combined.final.csv")
EEF2=subset(Q, Target == "EEF2")
IGFs=subset(Q, Target== "IGF1" | Target== "IGF2")
IGF1=subset(IGFs, Target == "IGF1")
IGF2=subset(IGFs, Target == "IGF2")

samplesize=summarySE(Q, measurevar="Copy_Number", groupvars=c("Sample", "Life_Stage"))
```

>With the intent to run linear models, the data was tested for the required assumptions. 

#Test of Assumptions for linear models
```{r}
#Statistical and visual test of heteroscedasticity
#All data
cn.test=lme(Copy_Number~Life_Stage, data = Q, na.action=na.omit, random=~1|Sample)
plot(Q$Copy_Number, residuals(cn.test))
abline(a=0,b=0)

#EEF2
eef2.cn.test=lme(Copy_Number~Life_Stage, data = EEF2, na.action=na.omit, random=~1|Sample)
#Residuals plot
plot(EEF2$Copy_Number, residuals(eef2.cn.test))
abline(a=0,b=0)
#Secondary Histogram to assess normality.
hist(residuals(eef2.cn.test))

#IGF1
igf1.cn.test=lme(Copy_Number~Life_Stage, data = IGF1, na.action=na.omit, random=~1|Sample)
#Residuals plot
plot(IGF1$Copy_Number, residuals(igf1.cn.test))
abline(a=0,b=0)
#Secondary Histogram to assess normality.
hist(residuals(igf1.cn.test))

#IGF2
igf2.cn.test=lme(Copy_Number~Life_Stage, data = IGF1, na.action=na.omit, random=~1|Sample)
#Residuals plot
plot(IGF2$Copy_Number, residuals(igf2.cn.test))
abline(a=0,b=0)
#Secondary Histogram to assess normality.
hist(residuals(igf2.cn.test))

```
>All genes fail the assumption for heteroscedasticity and normality. Data will be transformed to meet assumptions of the linear mixed model. In order to identify the best normalization method for our data, the bestNormalize package was used. Those results can be seen below. 

##Identification of possible transformation models by BestNormalize
```{r}
#analysis of whole data set (IGFs + EEF2)
bestNormalize(Q$Copy_Number)
#analysis of EEF2 alone
bestNormalize(EEF2$Copy_Number)
#analysis of IGFs
bestNormalize(IGFs$Copy_Number)
```

>For all three tests, the ordered quantile transformation was indicated as the most appropriate transformation method.


#DETERMINE THE EFFICIENCY OF EEF2 HAS A HOUSEKEEPING GENE
##Ordered Quantile transformation of all data
```{r}
orq=orderNorm(EEF2$Copy_Number)
EEF2=cbind(EEF2, orq=orq$x.t)

#Transformed Data Evaluation
model.orq.eef2=(lme(orq~Life_Stage, random =  ~1|Sample, data=EEF2))

plot(EEF2$orq, residuals(model.orq.eef2))
abline(a=0,b=0)

hist(residuals(model.orq.eef2))

#ORQ
##Run a lme to determine if there is significant variation in expression of the referene gene across life stages. The signficinat p values indicate that expression of the gene varies between sample stages, and is therefore not a suitable normalizing gene. 
post.test.orq=lme(orq~Life_Stage, data = EEF2, na.action=na.omit, random=~1|Sample)
anova(post.test.orq)

#Run a post-hoc test in the standard Tukey setup to get adjust p-values for pairwise comparison between groups
summary(glht(post.test.orq, linfct=mcp(Life_Stage="Tukey")))

#variance test
var=(aov(orq~Life_Stage, data=EEF2))
summary(var)
anova(var)
TukeyHSD(var)

EEF2.plot= ggplot(EEF2, aes(x=as.factor(Life_Stage), y=orq)) + 
  #this line creates a violin plot, not trimming the shape to hide data density on ends, and setting the width of the violins     to line up with the groups on the x axis.  
  geom_violin(trim=F, position= dodge, scale="width") +
  #A boxplot is overlayed on the violin plot to show the mean and quartiles of the data, and all dataset outlies are        highlighted in red.
  geom_boxplot(width=0.1, position= dodge, outlier.color = "red") +
  #the data points at each life stage are then placed on top of the boxplots to show the data distribution
  geom_point(position=dodge) +
  #change the y-axis to scientific notation 
  scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
  #Rename the x axis to display the developmental stages in "mid, early, and late" rather than numberical stages
  scale_x_discrete(name ="Life Stage", 
                   labels=c("Early-Dev", "Mid-Dev", "Late-Dev", "Day 0", "Week 4", "Week 8", "Adult")) +
  #change the y axis title to remove the underscore in coded dataframe
  ylab("Copy Number ORQ Transformed") +
  #bold the text on the axis titles
  theme(axis.title = element_text(face="bold", size=16)) +
  theme(axis.text = element_text(face="bold")) +
  theme(axis.text = element_text(size=14)) +
  theme(plot.title = element_text(face="bold", size=16)) +
  #remove legend from the plot
  theme(legend.position="none") +
  #add a title to the plot 
  ggtitle("EEF2 Not Valid Ref Gene (ORQ)")


ggsave(EEF2.plot, file="EEF2.norm.trans.png", width=12, height=10, dpi=300)

EEF2.plot
```

>The above shows that EEF2 has significant variance across time-points, making it not suitable for normalization. Therefore, data will not be normalized by this reference gene. 

#Anlaysis of IGF1 and IGF2

##Transformation of normalized IGF1 and IGF2 values
```{r}
#Veryify that after normalization, ORQ is still an appropriate transformation approach
bestNormalize(IGFs$Copy_Number)

#Obtained transformed values and bind to original data set
orq=orderNorm(IGFs$Copy_Number)
IGFs=cbind(IGFs, orq=orq$x.t)

#IGF1  
#Transformed Data Evaluation
igf1.or=subset(IGFs, Target == "IGF1")
igf1an.or=subset(igf1.or, Life_Status =="Animal")
orq.tr.igf1=(lme(orq~Life_Stage, random =  ~1|Sample, data=igf1an.or))
plot(igf1an.or$orq, residuals(orq.tr.igf1))
abline(a=0,b=0)
hist(residuals(orq.tr.igf1))


#IGF2  
#Transformed Data Evaluation
igf2.or=subset(IGFs, Target == "IGF2")
igf2an.or=subset(igf2.or, Life_Status =="Animal")
orq.tr.igf2=(lme(orq~Life_Stage, random =  ~1|Sample, data=igf2an.or))
plot(igf2an.or$orq, residuals(orq.tr.igf2))
abline(a=0,b=0)
hist(residuals(orq.tr.igf2))
```

>Ordered quantile transformation is still indicated as the best transformation method for this data set. Values were transformed and verified visually for meeting the assumptions of homoscedasticity and normality for linear models. Transformed values (which are previously normalized by EEF2) were added to original data set, and used in all further analyses.


#Determination of Statistical model using normalized and transformed data
```{r}
animal=subset(IGFs, Life_Status == "Animal")

model.igfs=(lme(orq~Life_Stage, random =  ~1|Sample, method="ML", data=animal))
summary(model.igfs)

model.igfs.sex=(lme(orq~Life_Stage+Sex, random =  ~1|Sample, na.action=na.omit, method="ML", data=animal))
summary(model.igfs.sex)
anova(model.igfs, model.igfs.sex)

```
>Sex is a non-significant piece of the model. When comparing the model with and witout sex as a fixed effect, adding sex does not significantly improve the model. The beta is also -0.02, compared to -1.45 to -1.25 in other variables. This is justification for removing Sex from the model.

##STATISTICAL MEASUREMENTS OF IGF1 AND IGF2 ACROSS LIFE STAGES
```{r}
#Test for effect of sex in model. A random command is added to account for the triplicate measures during qPCR analysis.
model.igfs.sex=anova(lme(orq~Target+Sex, random =  ~1|Sample, data=IGFs, na=na.omit)) 
model.igfs.sex

#Test if relative expression of IGF1 and IGF2 is significant between life stages across the entire data set.
model.igfs=summary(lme(orq~Target, random =  ~1|Sample, data=IGFs)) 
model.igfs
```

##SUBSETTING OF DATA BY LIFE STAGE. Relative IGF measures at each life stage. From this point on, data must be subsetted because the embryonic RNA extractions were whole organism, and the other samples were liver extractions. The two data types are not comparable.
```{r}
##Subset the IGF data set to seperate each life stage to run IGF1 vs IGF2 expression at each life stage. The whole model was significant, but this will determine whether or not IGF2 is higher at each life stage individually, or just over the entire model.
st2=subset(IGFs, Life_Stage == "Stage 2")
st6=subset(IGFs, Life_Stage == "Stage 6")
st16=subset(IGFs, Life_Stage == "Stage 16")
d0=subset(IGFs, Life_Stage == "Day 0")
w4=subset(IGFs, Life_Stage == "Week 4")
w8=subset(IGFs, Life_Stage == "Week 8")
adult=subset(IGFs, Life_Stage == "Adult")

##linear models are run comparing the expression levels (copy number) of IGF1 and IGF2 (target) at each life stage (data). A random command is added to account for the triplicate measures during qPCR analysis.
summary(lme(orq~Target, data=st2, random = ~1|Sample))
summary(lme(orq~Target, data=st6, random = ~1|Sample)) 
summary(lme(orq~Target, data=st16, random = ~1|Sample)) 
summary(lme(orq~Target, data=d0, random = ~1|Sample)) 
summary(lme(orq~Target, data=w4, random = ~1|Sample))
summary(lme(orq~Target, data=w8, random = ~1|Sample))
summary(lme(orq~Target, data=adult, random = ~1|Sample))
```

##SUBSETTING OF DATA BY LIFE STATUS (EMBRYO VS ANIMAL VS JUVENILE). IGF1 and IGF2 expression across development.
```{r}
##The IGF data set is now subsetted by life status, a less strict form of life stage where the samples are split between emrbyonic, animals (juveniles + adults), and juveniles alone. This is run to determine whether or not IGF2 expression is higher than IGF1 based on the main categories of development. Within each of these subcategories, the life status data is split to seperate IGF1 and IGF2 into seperate data frames, allowing for pairwise comparisons between life stages of development across time rather than comparing relative expression levels of IGF1 and IGF2.
Emb=subset(IGFs, Life_Status == "Embryo")
Emb1=subset(Emb, Target == "IGF1")
Emb2=subset(Emb, Target == "IGF2")

Animal=subset(IGFs, Life_Status == 'Animal')
Animal1=subset(Animal, Target == "IGF1")
Animal2=subset(Animal, Target == "IGF2")

Juv=subset(Animal, Life_Stage != "Adult")
Juv1=subset(Juv, Target == "IGF1")
Juv2=subset(Juv, Target == "IGF2")

##linear models are run comparing the expression levels (copy number) of IGF1 OR IGF2 (target) between each life stage across time/development. A random command is added to account for the triplicate measures during qPCR analysis.
#Is IGF1 expression statistically different between stage 16 and the other two stages?
emb1=(lme(orq~Life_Stage, random = ~1|Sample, data=Emb1))
anova(emb1)
emmeans(emb1, list(pairwise ~ Life_Stage), adjust = "tukey")

#rerun the above IGF1 embryonic samples on the undiluted cDNA rerun samples 
undil.emb=read.csv("undiluted.samples.csv")

orq.un=orderNorm(undil.emb$Copy_Number)
undil.emb=cbind(undil.emb, orq=orq.un$x.t)

emb1b=(lme(orq~Biological_Set, random = ~1|Sample, data=undil.emb))
anova(emb1b)
emmeans(emb1b, list(pairwise ~ Biological_Set), adjust = "tukey")

#Is IGF2 expression statistically different between stage 16 and the other two stages?
emb2=(lme(orq~Life_Stage, random = ~1|Sample, data=Emb2))
anova(emb2)
emmeans(emb2, list(pairwise ~ Life_Stage), adjust = "tukey")

#Is IGF1 expression statistically different between Day 0 and the other juvenile stages?
juv1=(lme(orq~Life_Stage, random = ~1|Sample, data=Juv1))
anova(juv1)
emmeans(juv1, list(pairwise ~ Life_Stage), adjust = "tukey")

#Is IGF2 expression statistically different between Day 0 and the other juvenile stages?
juv2=(lme(orq~Life_Stage, random = ~1|Sample, data=Juv2))
anova(juv2)
emmeans(juv2, list(pairwise ~ Life_Stage), adjust = "tukey")

#Is IGF1 expression statistically different between adulthood and the other juvenile stages?
anim1=(lme(orq~Life_Stage, random = ~1|Sample, data=Animal1))
anova(anim1)
emmeans(anim1, list(pairwise ~ Life_Stage), adjust = "tukey")

#Is IGF2 expression statistically different between adulthood and the other juvenile stages?
anim2=(lme(orq~Life_Stage, random = ~1|Sample, data=Animal2))
anova(anim2)
emmeans(anim2, list(pairwise ~ Life_Stage), adjust = "tukey")
```

##Heatmap
```{r}
#Read in data set that includes the percent of individuals expressing each BP in each tissue type
#this is from the presence/absense PCR data
heat=read.csv("heatmap.data2.csv")
#Create a heatmap showing the varying amount of expression between ages and tissue types across genes
map=ggplot(data= heat, aes(x=as.factor(Stage), y=Gene)) +
geom_tile(aes(fill=Expression), colour="white") + scale_fill_gradient(low="white", high="grey24")+
  theme(axis.text.y= element_text(size=(8), face="bold"),
        legend.text = element_text(size = 8, face="bold"),
        legend.title = element_text(size=8, face="bold")) +
   guides(fill=guide_legend(title="% Individuals"))+
  scale_x_discrete(name = "Tissue", labels=c("Early-Dev", "Mid-Dev", "Late-Dev", "Brain", "Gonad", "Heart", "Liver", "Muscle", "Tail", "Reg. Tail")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face="bold", size=8))

ggsave(map, file="heatmap.expression.png", width=5, height=2, dpi=300)

map
```

##Visualize DATA AS A WHOLE SET with raw data
```{r}
##NOTE: all graphs are back transformed to normalized copy number for easy comparison and evaluation.

##Graph IGF1 and IGF2 expression across all life stages. Notice Life_Stage_Code is used to reorder the samples by age rather than the default alphebetized order in gpplot.
all.ages= ggplot(IGFs, aes(x=as.factor(Life_Stage_Code), y=Copy_Number, fill= Target)) + 
  #this line creates a violin plot, not trimming the shape to hide data density on ends, and setting the width of the violins     to line up with the groups on the x axis.
  geom_violin(trim=F, position= dodge, scale="width") +
  #change the colors of IGF1 and IGF2 in the violin plot to white and grey fills respectively
  scale_fill_manual(values = c("white", "grey")) +
  #A boxplot is overlayed on the violin plot to show the mean and quartiles of the data, and all dataset outlies are          highlighted in red.
  geom_boxplot(width=0.1, position= dodge, outlier.shape = NA) +
  #Rename the x axis to display the developmental stages in "mid, early, and late" rather than numberical stages
  scale_x_discrete(name ="Life Stage", 
                   labels=c("Early-Dev", "Mid-Dev", "Late-Dev", "Day 0", "Week 4", "Week 8", "Adult")) +
  #Change theme elements of the text used on the axis to remove the axis and plot titles, remove the legend, and bold the axis    text
  theme(axis.title.x = element_blank()) +
  theme(axis.title.y  = element_blank()) +
  theme(plot.title = element_blank()) +
  theme(legend.position="none") +
  theme(axis.text = element_text(face="bold") +
  theme(axis.text = element_text(size= 14))) 

all.ages

##subset the IGF data to include only adult sames to visualize sex differences.
adult.sex=subset(IGFs, Life_Stage == "Adult")

##Plot the adult samples only, including data points colored by sex data
ad.sex= ggplot(adult.sex, aes(x=as.factor(Life_Stage_Code), y=Copy_Number, fill= Target)) + 
    geom_violin(trim=F, position= dodge, scale="width") +
    scale_fill_manual(values = c("white", "grey")) +
  geom_boxplot(width=0.1, position= dodge, outlier.shape = NA) +
  geom_point(position= dodge, aes(group= Target, colour=Sex)) +
    scale_color_manual(values=c("purple", "darkturquoise","clear")) +
  labs(colour = "Sex") +
  scale_x_discrete(name ="Life Stage", labels=c("Adult")) +
  theme(axis.title.y = element_text(face="bold")) +
  theme(legend.title = element_text(face= "bold")) +
  theme(axis.title.x = element_blank()) +
    theme(axis.title.y  = element_blank()) +
    theme(plot.title = element_blank()) +
  theme(legend.position="none") +
    theme(axis.text = element_text(face="bold") +
  theme(axis.text = element_text(size= 14))) 

ad.sex 

##Subset IGF data set to include none of the adult samples for plotting purposes. The adult expression of both IGF1 and IGF2 is much higher than the expression at any other stage, making it difficult to visualize. The plot is recreated witout the adult samples to compare expression across development and juvenile growth more closely. 
IGF_NA=subset(IGFs, Life_Stage != "Adult")

no.ad= ggplot(IGF_NA, aes(x=as.factor(Life_Stage_Code), y=Copy_Number, fill= Target)) + 
    geom_violin(trim=F, position= dodge, scale="width") + 
    scale_fill_manual(values = c("white", "grey")) +
    geom_point(position=dodge, aes(group= Target, colour=Sex)) +
  ##The na.translate tells it to include data points for the individuals with no sex, but to not annotate them in the legend. This can act up, so a second option is to create a blank label name in the legend for the NA points
  scale_color_manual(na.translate=FALSE, labels=c("Female", "Male", ""), values=c("purple", "darkturquoise","black")) +
  labs(colour = "Sex") +
   geom_boxplot(width=0.1, position= dodge, outlier.shape = NA) +
  scale_x_discrete(name ="Life Stage", 
                   labels=c("Early-Dev", "Mid-Dev", "Late-Dev", "Day 0", "Week 4", "Week 8")) +
  scale_y_continuous(labels = function(x) format(x, scientific = TRUE), limits = c(0 , 65000)) +
  theme(axis.title = element_text(face="bold")) +
  theme(plot.title = element_blank()) +
  theme(legend.title = element_text(face= "bold")) +
  theme(axis.title.y  = element_blank()) +
    theme(axis.title.x  = element_blank())+
  theme(axis.text = element_text(face="bold")+
  theme(axis.text = element_text(size= 14))) 

no.ad
```


##Export image for Journal
```{r}
##Load the cowplot package which will allow you to export the images in pre-determined proporations as one figure.
library("cowplot")

p=ggdraw() +
  #Export the first graph including all life stages. This graph will be on the top row, aligned along the left, take up 3/10 of the vertical space (beginning at 0.7/1 on the y-axis)+, and take up 6/10 of the horizontal space
  draw_plot(all.ages, x = 0, y = .7, width = .6, height = .3) +
  #Export the second graph including only the adult life stages. This graph will be on the top row, aligned to the right, and take up the reminaing 4/10 of the horizontal space on the top row. 
  draw_plot(ad.sex, x = 0.6, y = .7, width = .3, height = .3) +
 #Export the third image, which is in the second row, and spans the entire length of the figure horizontally. Vertically it takes up the first 0.3/10 of the figure beginning at 0 on the y-axis.
  draw_plot(no.ad, x = 0, y = 0, width = 1, height = 0.7) +
 #Add labels to each of the figures. They will be labeled in the order that they are exported. The text size and exact coordiate placement is shown with the x and y values. 
   draw_plot_label(label = c("A", "B","C"), size = 15,
                   #Note the y values are slightly above 1 on the top row, to avoid printing on top of                      the axis labels.
                  x = c(0, 0.6,0), y = c(1.02, 1.02,0.72)) 
  

# Add additional margin space to the edge of the graph to allow for formatting and editing of axis and graph titles. 
p2= p + theme(plot.margin=unit(c(15,15,15,15),"mm")) 

##Save the image as a png file with predefined dimensions. Add the dpi value to export it as publication quality. 
ggsave(p2, file="publication.image.png", width=12, height=10, dpi=300)

p2
```


#Fold change calculation
```{r}
adult.comp<- summarySE(adult.sex, measurevar="oqr", groupvars=c("Target"))
adult.comp

animal.comp<- summarySE(Animal, measurevar="orq", groupvars=c("Target", "Life_Stage"))
animal.comp
```
